---
title: "Life Expectancy EDA"
author: "Adina Zhang"
date: "April 3, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)

# Load dataset
life_exp = read_csv("./Life Expectancy Data.csv") %>% janitor::clean_names()
```


## EDA

```{r}
summary = skimr::skim_to_list(life_exp)
rbind(summary$integer, summary$numeric) %>% 
  select(variable:p100) %>% 
  knitr::kable()


# Continuous variables
cont = life_exp %>% select(-country, -status)
# Scatterplots 
pairs(cont[,2:6])
pairs(cont[,c(2,7:10)])
pairs(cont[,c(2,11:14)])
pairs(cont[,c(2,15:17)])
pairs(cont[,c(2,18:20)])

# Binary variable
life_exp %>% ggplot(aes(y = life_expectancy, color = status)) + 
  geom_boxplot()
```

Non-linear variables:
* log gdp
* hiv_aids
* gdp
* alcohol
* bmi
* thinness_5_9_years
* hepatitis_b
* total_expenditure
* thinness_1_19_years
* thinness_5_9_years
* log transformed under_five_deaths

We should eliminate a few variables to create a slightly more parsimonious model.

Variables that we will eliminate:
* population (no significant relationship to life expectancy)
* infant deaths (highly correlated with under five deaths)
* measles (no significant relationship to life expectancy)
* polio or diptheria (highly correlated to each other)
* percentage_expenditure (highly correlated to loggdp)

```{r}
# Dataset for analysis
life_analysis = life_exp %>% 
  select(-country, -population, -infant_deaths, -measles, -polio,
         -percentage_expenditure) %>% 
  mutate(log_gdp = log(gdp + 0.1),
         log_under_five = log(under_five_deaths + 0.1)) %>% 
  select(-gdp, -under_five_deaths)

```

