---
title: "Life Expectancy EDA"
author: "Adina Zhang"
date: "April 3, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)

# Load dataset
life_exp = read_csv("./Life Expectancy Data.csv") %>% janitor::clean_names()
```


## EDA

```{r}
# Continuous variables
cont = life_exp %>% select(-country, -status)
# Scatterplots 
pairs(cont[,2:6])
pairs(cont[,c(2,7:10)])
pairs(cont[,c(2,11:14)])
pairs(cont[,c(2,15:17)])
pairs(cont[,c(2,18:20)])
```

Non-linear variables:
* percentage_expenditure
* hiv_aids
* gdp
* alcohol
* bmi
* thinness_5_9_years

We should eliminate a few variables to potentially avoid multicollinearity. Check collinearity between gdp, percentage expenditure, and total expenditure. When looking at some of the correlation tables, there are sometimes two distinct slopes. This probably comes from the "developing" class. 

Variables that show very little correlation with life expectancy:
* population
* measles
* hepatitis b
* total expenditure
* thinness_1_19_years